%%
%% File: cache.tex
%% Author: Octavian Procopiuc <tavi@cs.duke.edu>
%% Created: Sep 26, 2000
%%
%% $Id: cache.tex,v 1.2 2001-12-19 21:18:30 tavi Exp $
%%

\chapter{The Cache Manager}

\section{Overview}
\label{cache:overview}

\section{Class Declaration}

   \begin{tabbing}
   \hspace*{.3in} \= \hspace{.5in} \= \\

   \> {\tt template<class T, class W> class AMI\_CACHE\_MANAGER;}
   \end{tabbing}

\section{Constructors and Destructor}
   \begin{tabbing}
   \hspace*{.3in} \= \hspace{.5in} \= \\

   \> {\tt AMI\_CACHE\_MANAGER(size\_t capacity)}\\ 
   \>\>\parbox[t]{5.5in}{Construct a fully-associative cache manager with the given capacity.}\\[3mm]

   \> {\tt AMI\_CACHE\_MANAGER(size\_t capacity, size\_t assoc)}\\ 
   \>\>\parbox[t]{5.5in}{Construct a cache manager with the given capacity and associativity.}\\[3mm]

   \> {\tt $\sim$AMI\_CACHE\_MANAGER()}\\ 
   \>\>\parbox[t]{5.5in}{Destructor. Write out all items still in the cache.}\\[3mm]

   \end{tabbing}

\section{Member functions}
   \begin{tabbing}
   \hspace*{.3in} \= \hspace{.5in} \= \\ 

   \> {\tt  bool read(size\_t k, T \& item);}\\ 
   \>\>\parbox[t]{5.5in}{Read an item from the cache based on key {\tt k} and store it in {\tt item}. If found, the item is removed from the cache. Return true if the key was found.}\\[3mm]

   \> {\tt  bool write(size\_t k, const T \& item);}\\ 
   \>\>\parbox[t]{5.5in}{Write an item in the cache based on the given key {\tt k}. If the cache was full, the least recently used item is writen out using the {\tt W} function object, and it is removed from the cache.}\\[3mm]

   \> {\tt  bool erase(size\_t k);}\\ 
   \>\>\parbox[t]{5.5in}{Erase an item from the cache based on the given key {\tt k}. Return true if the key was found.}\\[3mm]
   \end{tabbing}

\section{Implementation Details}

{\tt AMI\_CACHE\_MANAGER} is a macro definition which resolves to {\tt
AMI\_cache\_manager\_lru}, an LRU cache manager, the only implementation.
The cache is an array of pairs (key, item of type {\tt T}). The array is
divided into (capacity/assoc) sets, and inside a set item are inserted in
position 0 and deleted from the last position. This insures LRU behavior.
