%%
%% Copyright (c) 1994, 1995 Darren Vengroff
%%
%% File: tpie.tex
%%
%% $Id: tpie.tex,v 1.11 1999-06-18 17:28:13 rajiv Exp $
%%

\documentstyle[11pt,html,makeidx]{book}

% normal margins for US-size paper
\setlength{\topmargin}{-.5in}   
\setlength{\oddsidemargin}{.1in} % distance from left edge of page to text
\setlength{\evensidemargin}{.1in} % distance from left edge of page to text
\setlength{\textwidth}{6.3in}
\setlength{\textheight}{9in}

\makeindex

% Additions to makeidx.sty
%\makeatletter
%\@ifundefined{alsoname}%
%   {\def\alsoname{also}}{}

%\def\seealso#1#2{{\em \seename\ \alsoname\/} #1}
%\makeatother

\newcommand{\edition}{0.9.01a}
\newcommand{\version}{0.9.01a}
\newcommand{\gxxversion}{2.8.1}

%% Commands to mimic some that latexinfo has:

%% End of LaTeXinfo-like commands.

\begin{document}

\title{TPIE User Manual and Reference}
\author{Lars Arge \and Rakesh Barve \and Octavian Procopiuc \and Laura
Toma\and 
        Darren Erik Vengroff \and Rajiv Wickeremesinghe.}
\date{{\bf DRAFT} of \today}

\maketitle

\begin{titlepage}

\vspace*{5 ex}

\centerline{\Large {\bf DRAFT} of \today}

\vspace{\fill}


\noindent TPIE User Manual and Reference

\noindent Edition \edition, for TPIE version \version.

\vspace{2 ex}

\noindent Copyright \copyright 1994, 1995 Darren Erik Vengroff

\vspace{2 ex}

The programs described in this manual are free software; you can
redistribute them and/or modify them under the terms of the GNU General
Public License as published by the Free Software Foundation; either
version 2 of the License, or (at your option) any later version.

This manual and the programs it describes are distributed in the hope
that they will be useful, but WITHOUT ANY WARRANTY; without even the
implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
PURPOSE.  See the GNU General Public License (Appendix~\ref{app:gpl})
for more details.

You should have received a copy of the GNU General Public License
along with this manual; if not, write to the Free Software Foundation,
Inc., 675 Mass Ave, Cambridge, MA 02139, USA.
\end{titlepage}

\tableofcontents

\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}

This manual describes TPIE, a Transparent Parallel I/O Environment,
designed to allow programmers to write high performance I/O-efficient
programs for a variety of platforms.

{\bf This manual, like the whole of the TPIE project, is a work in
  progress.}  The author is making it available in in its current state
in the hopes that it will be useful, but without any warrantee
whatsoever.  Please refer to the copyright page at the beginning of
this manual for full details.

\part{User Manual}
  \input{user_manual}

\part{Reference}
  \input{ami_reference} % Chapter: AMI Programmers' Reference
  \input{implementation} % Chapter: The Implementation of TPIE
  \input{tuning} % Chapter: TPIE Performance Tuning
  \input{applications} %  Chapter: Test and Sample Applications
  \input{acknowledgements} % Chapter: Acknowledgements

\bibliographystyle{abbrv}
\bibliography{tpie}

\part{Appendices}
  \input{appendices} 

\end{document}


