%%
%% $Id: tuning.tex,v 1.7 1999-11-17 20:56:55 tavi Exp $
%%
\chapter{Configuration and Performance Tuning}
\label{sec:tuning}\index{Configuration}\label{sec:configuration}

\comment{LA: Should there be something about compiling in this chapter (and
about e.g including ami.h)?}

\section{TPIE Configuration}
Certain behaviours of TPIE at run-time are controlled by
compile-time variables, whose values should be defined
before including any TPIE headers. Depending on the options
desired, the values of these
variables can be specified as early as when TPIE is
installed, or as late as when an individual application
program is compiled. Section \ref{sec:customization}
described the options available at installation
time. Section \ref{sec:appconfig} describes how TPIE can be
configured differently for individual
TPIE applications.

\subsection{Installation Options}
\index{Customization} \label{sec:customization}

\comment{LA: We need to change this when we change logging}
It is possible to customize the TPIE by providing
arguments to the {\tt configure}
script when TPIE is first installed (see Section \ref{sec:installation}.\index{configuration:options} None of these
 arguments are necessary and the first time you build TPIE
 you should not need any of them. The arguments
recognized are as follows:
\begin{description}
\item[\myverb{--enable-log-lib}] 
  \index{enable-log-lib@{\tt --enable-log-lib}}
  Enable logging in TPIE library code.
  This can also be accomplished at compile time by defining the macro
  \myverb{TP\_LOG\_LIB} using the syntax \myverb{make lib TP\_LOG\_LIB=1}.
  This is useful for debugging the TPIE library, but slows it down.
  This option works by defining \myverb{TPL\_LOGGING}
  \index{TPL_LOGGING@{\tt TPL\_LOGGING}}
  when compiling the library. 
  Section \ref{sec:logging} discusses TPIE logging.
\item[\myverb{--enable-assert-lib}]  
  \index{enable-assert-lib@{\tt --enable-assert-lib}}
  Enable assertions in the TPIE library code for debugging purposes.
  This can also be accomplished at compile time by defining the macro
  \myverb{TP\_ASSERT\_LIB} using the syntax \myverb{make lib TP\_ASSERT\_LIB=1}.
  This option works by defining \myverb{DEBUG\_ASSERTIONS}
  \index{DEBUG_ASSERTIONS@{\tt DEBUG\_ASSERTIONS}} 
  when compiling the library.
\item[\myverb{--enable-log-apps}]  and
\item[\myverb{--enable-assert-apps}]  
  \index{enable-assert-apps@{\tt --enable-assert-apps}}
  \index{enable-log-apps@{\tt --enable-log-apps}}
  Similar to {\tt --enable-log-lib} and {\tt --enable-assert-lib}, but
  they apply to the test application code.  Running \myverb{make test}
  with the options \myverb{TP\_LOG\_APPS=1} and/or \myverb{TP\_ASSERT\_APPS=1}
  accomplishes the same thing.
\item[\myverb{--enable-expand-ami-scan}]  Expand the macros in the file
  \myverb{ami\_scan.h} when making the include directory with the
command {\tt make include} (or {\tt make all}).  This is mainly useful for
debugging the code in \myverb{ami\_scan.h} itself, and is not normally
needed by TPIE programmers.  It may make compilation of TPIE programs
slightly faster because the macro processor of the C++ compiler will
have less work to do.  In addition to the standard GNU tools mentioned
in Section~\ref{sec:gnu-software}, this requires \myverb{perl}.
\item[\myverb{--disable-*}]  Any of the options above can be explicitly
  disabled  by using this syntax.  For example
  \myverb{--disable-expand-ami-scan}. 
\end{description}

\subsection{Configuring TPIE for Individual Applications}\label{sec:appconfig}

Certain TPIE configuration options can be selected by
setting compile-time variables in the file
\myverb{app\_config.h}\index{app_config.h@{\tt app\_config.h}}
which is then included in an application program. A typical
example of this file can be found in the \myverb{test}
directory. Selected parts of the file are shown
and discussed below. \comment{LA: Something general about TPIE
   configuration as set up by the configure-script needs to
   be included here (e.g. discuss config.h file).}

\subsection{app\_config.h}

\begin{verbatim}
// Get the configuration as set up by the TPIE configure script.
#include <config.h>

/* ********************************************************************** */
/*                      developer use                                     */
/* ********************************************************************** */


/* ********************************************************************** */
/*                       choose BTE                                       */
/* ********************************************************************** */

/* Pick a version of BTE streams; default is BTE_IMP_UFS */
//#define BTE_IMP_MMB
//#define BTE_IMP_STDIO
#define BTE_IMP_UFS


/* ********************************************************************** */
/*                      configure BTE                                     */
/* ********************************************************************** */


/* ********************************************************************** */
/* BTE_MMB configuration options */
/* ********************************************************************** */
#ifdef BTE_IMP_MMB

/* define logical blocksize; default is 32 * operating system blocksize */
#define BTE_MMB_LOGICAL_BLOCKSIZE_FACTOR 32

/* enable/disable TPIE read ahead; default is enabled (set to 1) */
#define BTE_MMB_READ_AHEAD 1

/* read ahead method, ignored unless BTE_MMB_READ_AHEAD is set to 1;
   if USE_LIBAIO is enabled, use asynchronous IO read ahead; otherwise
   use use mmap-based read ahead; default is mmap-based read ahead
   (USE_LIBAIO not defined) */
//#define USE_LIBAIO

#endif


/* ********************************************************************** */
/* BTE_UFS configuration options */
/* ********************************************************************** */
#ifdef BTE_IMP_UFS

/* define logical blocksize; default is 32 * operating system blocksize */
#define BTE_UFS_LOGICAL_BLOCKSIZE_FACTOR 32

/* enable/disable TPIE read ahead; default is disabled (set to 0) */
#define BTE_UFS_READ_AHEAD 0

/* read ahead method, ignored unless BTE_UFS_READ_AHEAD is set to 1;
   if USE_LIBAIO is set to 1, use asynchronous IO read ahead;
   otherwise no TPIE read ahead is done; default is disabled (set to
   0) */
#define USE_LIBAIO 0

#endif
/********************************************************************/


/********************************************************************/
/*  THE FOLLOWING MACROS ARE NORMALLY NOT MODIFIED BY USER           */
/********************************************************************/

/* Use the single BTE stream version of AMI streams; in the current
   option this is the only option */
#define AMI_IMP_SINGLE

/* enable/disable virtual interface; normally disabled */
#ifndef AMI_VIRTUAL_BASE
#define AMI_VIRTUAL_BASE 0
#endif
#ifndef BTE_VIRTUAL_BASE 
#define BTE_VIRTUAL_BASE 0
#endif


/********************************************************************/
/*                            logging;                              */
/*              this should NOT be modified by user!!!              */
/*       in order to enable/disable library/application logging,    */
/*     run tpie configure script with appropriate options           */
/********************************************************************/
// Use logs if requested.
#if TP_LOG_APPS
#define TPL_LOGGING 1
#endif

#include <tpie_log.h>

// Enable assertions if requested.
#if TP_ASSERT_APPS
#define DEBUG_ASSERTIONS 1
#define DEBUG_CERR 1
#define DEBUG_STR 1
#endif
#include <tpie_assert.h>
/********************************************************************/
\end{verbatim}

\subsection{Macros in app\_config.h}

\begin{description}
\item[{\myverb{BTE\_IMP\_*}}] Used to choose which of the available Block
Transfer Engine (see Section~\ref{sec:ref-bte}) implementations to
use. Version \version~of TPIE is distributed with three BTEs and the
desired BTE is chosen by defining \myverb{BTE\_IMP\_STDIO}, \myverb{BTE\_IMP\_MMB}
or \myverb{BTE\_IMP\_UFS}. See Section~\ref{sec:ref-bte} for a discussion of
the implementation details in these BTEs. The next section discusses how to
choose an appropriate BTE for a given application in order to obtain
maximal performance.

  \index{BTE_IMP_*@{\tt BTE\_IMP\_*}}
  \index{block transfer engine!implementation}
  \index{implementation!BTE}
\end{description}

\noindent
If \myverb{BTE\_IMP\_MMB} or \myverb{BTE\_IMP\_UFS} is defined, the following macros
are used to control BTE options (how to set the options for maximal
performance is discussed in the next section):

  \begin{description}

  \item[{\myverb{BTE\_\*\_LOGICAL\_BLOCKSIZE\_FACTOR}}] This macro sets the
  logical blocksize used by the BTE in units of the physical block size
  (refer to Section~\ref{sec:ref-bte}). Value 1 indicates that the logical
  blocksize is the same as the physical blocksize of the OS.

  \item[{\myverb{BTE\_*\_READ\_AHEAD}}] Defining this macro instructs the
  BTE to optimize for sequential read speed by reading blocks into main
  memory before the data they contain is actually needed.

  \index{BTE_MMB_READ_AHEAD@{\tt BTE\_MMB\_READ\_AHEAD}}
  \index{read ahead}
  
  \item[{\myverb{USE\_LIBAIO}}] If \myverb{BTE\_MMB\_READ\_AHEAD} is defined,
  defining this macro results in the read ahead being performed using the
  asynchronous I/O library \myverb{libaio}. If the macro {\myverb{USE\_LIBAIO}}
  is not defined the read ahead is done using \myverb{mmap} and double
  buffering in the case of \myverb{BTE\_IMP\_MMB} and not done at all in the
  case of \myverb{BTE\_IMP\_UFS} (refer to Section~\ref{sec:ref-bte}).

  \index{libaio library@{\tt libaio}}
  \index{USE\_LIBAIO }
  \end{description}

\noindent
The rest of the macros are normally not modified by TPIE application
programmer:

\begin{description}
\item[{\myverb{AMI\_IMP\_*}}] This macro controls which Access Method
Interface implementations (see Section~\ref{sec:ref-bte}) to use. Version
\version~of TPIE is only distributed with one AMI implementation, which stores
the contents of a given stream on a single disk. This implementation is
selected by defining \myverb{AMI\_IMP\_SINGLE}.

  \index{AMI_IMP_*@{\tt AMI\_IMP\_*}}
  \index{access method interface!implementation}
  \index{implementation!AMI}
  \index{implementation!AMI!single disk}

\item[{\myverb{AMI\_VIRTUAL\_BASE}}] Defining this macro makes the base class
declares a large number of virtual functions for the class
\myverb{AMI\_base\_stream}, which is the base class of all implementations of
AMI streams. This is useful for debugging new AMI stream implementations,
but many compilers cannot properly inline virtual functions, so it slows
the system down significantly.

  \index{AMI_VIRTUAL_BASE@{\tt AMI\_VIRTUAL\_BASE}}
  \index{virtual base class!AMI} 

\item[{\myverb{BTE\_VIRTUAL\_BASE}}] Similar to \myverb{AMI\_VIRTUAL\_BASE}, but
for the BTE layer.
  
  \index{BTE_VIRTUAL_BASE@{\tt BTE\_VIRTUAL\_BASE}}
  \index{virtual base class!BTE} 

%All the AMI streams in TPIE are instances of a class derived from the base
%stream class \myverb{AMI\_base\_stream}. Although this base class has no
%functionality in itself, it can provide the minimum set of methods that
%need to be implemented by a derived class in order to have the required
%stream functionality. This is done by using abstract virtual methods. To
%tell TPIE to declare these methods in the base stream class you need to
%set \myverb{AMI\_VIRTUAL\_BASE} to 1 (either in the application's
%configuration, or during compilation by using the -D flag). Similarly,
%\myverb{BTE\_VIRTUAL\_BASE} controls the behavior of the BTE streams.
%
%By default, both BTE and AMI streams use a non-virtual base. By doing this,
%the performance of the stream methods is improved dramatically, in many
%cases. The biggest gain is obtained from inlining short methods (virtual
%methods cannot be inlined). 
%
%We believe this setup provides a good compromise between the need for high
%performance one one hand, and clarity, modularity and debugging support on
%the other hand.


\item[{\myverb{TPL\_LOGGING}}] Set to a non-zero value to enable logging of
TPIE's internal behavior.\comment{LA: Is this correct?} By default,
information is logged to the log file\index{log file} \myverb{/tmp/TPLOG\_XXX}
where \myverb{XXX} is a unique system dependent identifier. Typically it
encodes the process ID of the TPIE process that produced it in some
way. See Section~\ref{sec:logging} for information on exactly what TPIE
writes to the log file.

  \index{TPL_LOGGING@{\tt TPL\_LOGGING}}

\item[{\myverb{DEBUG\_ASSERTIONS}}] Define to enable TPIE assertions. These
assertions check for inconsistent or erroneous conditions within TPIE
itself. They are primarily intended to aid in the debugging of TPIE. Some
overhead is added to programs compiled with this macro set.

  \index{DEBUG_ASSERTIONS@{\tt DEBUG\_ASSERTIONS}}
  \index{debugging!TPIE}

\item[{\myverb{DEBUG\_CERR}}] Defining this macro tells TPIE to write all
internal assertion messages to the C++ standard error stream \myverb{cerr} in
addition to the TPIE log file.

 \index{DEBUG_CERR@{\tt DEBUG\_CERR}}
 \index{debugging!TPIE}

\item[{\myverb{DEBUG\_STR}}] Defining this macro enables certain debugging
messages that report on the internal behavior of TPIE but do not
necessarily indicate error conditions. In some cases this can increase the
size of the log dramatically.

  \index{DEBUG_STR@{\tt DEBUG\_STR}}
  \index{debugging!TPIE}

\end{description}


\subsection{Environment Variables}
\index{Environment variables}\label{sec:environment}

In version \version~of TPIE there is only one environment variable. The
variable is called \myverb{AMI\_SINGLE\_DEVICE} and defines where TPIE places
temporary streams. The default location is \myverb{/var/tmp}. If a different
location is desired, \myverb{AMI\_SINGLE\_DEVICE} must be set accordingly. For
example (in C-shell): setenv AMI\_SINGLE\_DEVICE /usr/project/tmp/.
\index{configuration|)}


\section{TPIE Performance Tuning}
\index{performance tuning}

\subsection{Choosing and Configuring a BTE Implementation}
\label{sec:choosingbte}

\comment{LA: This needs an overhaul!}

Choosing an appropriate BTE implementation (and BTE parameter settings) for
best performance is both application and system dependent. (See
section~\ref{sec:ref-bte} for a description of the three BTEs TPIE is
currently distributed with). Theoretically, \myverb{BTE\_mmb} should have the
best performance for most applications, because space and copy time is
saved relative to \myverb{BTE\_stdio} and \myverb{BTE\_ufs} as steam objects do
not have to pass through kernel level buffer space when accessed. On the
other hand, buffering and prefetching has to be explicitly implemented in
\myverb{BTE\_mmb} whereas it is (typically) done by the OS in \myverb{BTE\_stdio}
and \myverb{BTE\_ufs}. Also theoretically, \myverb{BTE\_ufs} (and \myverb{BTE\_mmb})
should perform better than \myverb{BTE\_stdio} because of fewer kernel calls
and because of the (possible) larger logical block size. However, in
practice the performance of the three BTE's are very system (and
application) dependent. This is for example due to different
implementations of the \myverb{fread()}, \myverb{fwrite()}, \myverb{read()},
\myverb{write()}, \myverb{mmap()}, and \myverb{munmap()} calls on different
machines.\comment{LA: Other reasons?}

The most important BTE configuration parameter is the logical block size
setting of the \myverb{BTE\_mmb} and \myverb{BTE\_ufs}. As discussed in
Section~\ref{sec:ref-bte}, a large setting increases performance due to the
less kernel calls and due to the (track) buffering and prefetching in the
disk controller, while a too large setting results in decreased performance
due to the BTE's use of main memory. Thus this parameter should be chosen
carefully. As far at the other BTE configuration parameters (prefetching)
are concerned, the default settings in the \myverb{app\_config.h} file in the
\myverb{test} directory are normally the best.

%One criterion that should be carefully considered while setting BTE
%parameters is the \myverb{BTE\_LOGICAL\_BLOCKSIZE\_FACTOR}. This parameter
%determines the unit of I/O and buffering used by the BTE stream
%implementation. In case of \myverb{BTE\_mmb} or \myverb{BTE\_ufs}
%implementations, the amount of memory dedicated to a stream is either
%\myverb{BTE\_LOGICAL\_BLOCKSIZE\_FACTOR} times the operating system blocksize
%(no prefetching) or twice \myverb{BTE\_LOGICAL\_BLOCKSIZE\_FACTOR} times the
%operating system blocksize (in the presence of prefetching). So the value
%of the \myverb{BTE\_LOGICAL\_BLOCKSIZE\_FACTOR} parameter, together with
%available memory, determines the number of BTE streams (and hence AMI
%streams) that can be active or ``alive'' at any time. This is an upper
%bound on the arity of a multi-way merge or a multi-way distribution
%operation that can be undertaken by a TPIE application; which can have a
%crucial impact on (say, the number of passes required in external sorting
%and hence the) net running time. The size of each buffer and the size of
%each I/O in the BTE stream is \myverb{BTE\_LOGICAL\_BLOCKSIZE\_FACTOR} times the
%operating system blocksize, so this roughly corresponds to the amount of
%data brought in or written out at the cost of a single disk operation.  The
%latter observation suggests that \myverb{BTE\_LOGICAL\_BLOCKSIZE\_FACTOR} should
%be set to a high value; but a high value for this parameter inhibits the
%number of streams active at a time and hence can result in an increase in
%the number of passes required in sorting.
%%In the case of external memory indexing data structures based on trees, the
%value of the \myverb{BTE\_LOGICAL\_BLOCKSIZE\_FACTOR} for any BTE stream (or BTE
%block collection, in future TPIE versions) used to implement the external
%memory data structure should be made as close to the size of the tree node
%as possible.

In order to help in deciding which BTE to choose for a given
application/system, as well as deciding on what logical block size to use
(in \myverb{BTE\_mmb} and \myverb{BTE\_ufs}), we have included a C program in the
\myverb{test} directory of the TPIE distribution called \myverb{bte\_test.c}.
This program can be used to determine the streaming speeds attained by
\myverb{BTE\_stdio}, \myverb{BTE\_mmb}, and \myverb{BTE\_ufs} streams on a given
system. The program simulates the buffering and I/O mechanisms used by each
of the BTE stream implementations so that the ``raw'' (in the sense that
there is no TPIE layer between the program and the filesystem) streaming
speed of an I/O-buffering mechanism combination can be determined. To use
the program, define one of \myverb{MMAP\_TEST},
\myverb{READ\_WRITE\_TEST}\comment{LA: Why not UFS\_TEST?} or \myverb{STDIO\_TEST}
in the program depending on whether you want to test the streaming speed of
\myverb{BTE\_mmb}, \myverb{BTE\_ufs} or \myverb{BTE\_stdio}. Also define the
\myverb{BLOCKSIZE\_BASE} parameter to be equal to the underlying operating
system blocksize.\comment{LA: Why not automatic?} Compile the program using
a C compiler. In order to test the streaming performance of BTE streams of
objects of size \myverb{ItemSize}, the program first writes out some
specified number \myverb{NumStreams} of BTE streams containing a specified
number \myverb{NumItems} of items. Then it carries out a perfect
\myverb{NumStreams}-way interleaving of the streams via a simple merge like
process, writing the output to an output stream. During the computation,
each of the \myverb{NumStreams} streams input to the merge, as well as the
stream being output by the merge uses either one (when
\myverb{READ\_WRITE\_TEST} or \myverb{STDIO\_TEST} are set to 1) or two (when
\myverb{MMAP\_TEST} is set to 1) buffers.  In case of \myverb{STDIO\_TEST}, the
buffers are not maintained in the program but by the stdio library. In the
case of \myverb{MMAP\_TEST} or \myverb{READ\_WRITE\_TEST}, each buffer is set to
be of size \myverb{block\_factor} times \myverb{BLOCKFACTOR\_BASE}, and each I/O
operation corresponds to a buffer-sized operation. To test the streaming
performance of a BTE stream with \myverb{items\_in\_block} items in each block
simply execute:
\begin{verbatim}
bte_test NumItems ItemSize NumStreams block_factor items_in_block DataFile 
\end{verbatim}
The output of the program (streaming speed) is appended to the file
\myverb{DataFile}. The streaming speed, alternatively called I/O Bandwidth,
is given in units of MB/s, and can be used to decide which BTE to use and
how to configure it.

\subsection{Other Factors Affecting Performance}

In addition to the choice (and configuration) of BTE, a number of other
factors, not all of which are TPIE specific, can effect the performance of
a TPIE application.

\begin{description}
\item[Inlining operation management object methods] Failing to inline the
\myverb{operate()} method of operation management objects can be a major
source of lackluster performance of an application, since \myverb{operate()}
is called once for every object in a stream being scanned. Inlining of
\myverb{operate()} is, of course, just a suggestion to the compiler, which
can choose to ignore it. In order to maximize the likelihood of inlining,
it is a good idea to keep the function  short and simple. One way of
doing this is to wrap complex pieces of code that are called less often in
separate functions.
\item[\myverb{gcc} optimization] We recommend using the \myverb{-O2} level of
optimization of \myverb{gcc} in order to obtain the best overall
performance. Although better performance can normally be obtained using
\myverb{-O3}, this optimization leads to increased program size which can
potentially result in decreased performance.
\item[Memory size] To insure that no disk swapping is done by the OS, the
size of main memory used by TPIE (set by \myverb{MM\_manager.resize\_heap()},
see Section\ref{sec:compiling} and Section~\ref{sec:mm-ref}) should be set
to a realistic value. The best value is usually much smaller than the size
of the memory installed in the computer (due to memory use of operating
system resources and daemons).
\end{description}

%\section{Using Multiple Physical Devices}


\section{TPIE Logging}\index{logging}
\label{sec:logging}

\comment{LA: The whole logging thing needs to be reevaluated and changed}
When logging is turned on (see Section \ref{sec:configuration}), TPIE
creates a log file\index{log file} in \myverb{/tmp/TPLOG\_XXXXXX}, where
\myverb{XXXXXX} is a unique system dependent identifier. TPIE writes into
this file using a \myverb{logstream} class, which is derived from
\myverb{ofstream} and has the additional functionality of setting a priority
and a threshold for logging. If the priority of a message is below
 the threshold, the message is not logged. There are four priority
levels defined in TPIE, as follows.
\begin{description}
\item[\myverb{TP\_LOG\_FATAL}] is the highest level and is used for all kinds
of errors that would normally impair subsequent computations. Errors are
always logged;
\item[\myverb{TP\_LOG\_WARNING}] is the next lowest and is used for warnings.
\item[\myverb{TP\_LOG\_APP\_DEBUG}] is used by applications built on top of TPIE, for logging debugging information.
\item[\myverb{TP\_LOG\_DEBUG\_INFO}] is the lowest level and is used by the TPIE library for logging debugging information.
\end{description}
By default, the threshold of the log is set to the lowest level, \myverb{TP\_LOG\_DEBUG\_INFO}. To change the threshold level, the following macro is provided:
\begin{quote}
\myverb{LOG\_SET\_THRESHOLD(}{\em level}\myverb{)}
\end{quote}
The threshold level can be reset as many times as needed in a program. This enables the developer to focus the debugging effort on a certain part of the program.

Four macros are provided for writing into the log:
\begin{quote}
\myverb{LOG\_FATAL}({\em msg})

\myverb{LOG\_WARNING}({\em msg})

\myverb{LOG\_APP\_DEBUG}({\em msg})

\myverb{LOG\_DEBUG\_INFO}({\em msg}),
\end{quote}
where {\em msg} is the information to be logged; {\em msg} can be any type
that is supported by the \myverb{C++} \myverb{fstream} class. Each of these
macros sets the corresponding priority and sends {\em msg} to the log
stream.

%{\em Logging should always be done using one of the above macros.} Any
%other method of logging could hinder the ability of TPIE to turn off
%logging and, as a result, could affect performance.


%\subsection{Template Instantiation}

%{\bf Important Note:} Much of the information in this section is
%likely to change as the template instantiation mechanism of the 
%{\tt g++}\index{g++@{\tt g++}} compiler improves.  If you are
%interested in the nitty gritty details of template instantiation,
%consult~\cite{ellis:arm} or one of the frequent discussions on the
%topic in the newsgroup {\tt comp.lang.c++}
%\index{comp.lang.c++@{\tt comp.lang.c++}}.

%\index{templates!instantiation|(}
%\noindent Most of the classes and functions TPIE defines are
%templated.  Furthermore, many user written operation management
%object\index{operation management objects!user supplied} classes are
%likely to be templated; many of those supplied with the test and
%sample applications are.

%Unfortunately, many C++\index{C++} compilers do not properly implement
%templated function and/or classes.  In particular, the GNU C++
%compiler, {\tt g++}\index{g++@{\tt g++}}, version \gxxversion, which
%was used in the development of TPIE has some deficiencies when it
%comes to template instantiation.  It also has a well defined mechanism
%for working around these deficiencies, which TPIE takes significant
%advantage of.  This mechanism prevents the compiler from implicitly
%instantiating any template.  Thus, all templates used by a program
%must be explicitly instantiated at compile time or they will not be
%available at link time and linking will fail.

%In order to tell {\tt g++}\index{g++@{\tt g++}} not to implicitly
%instantiate any templates, the {\tt -fno-implicit-templates} flag is
%used.  Additionally, the macro {\tt NO\_IMPLICIT\_TEMPLATES} should be
%defined on the command line, using {\tt -D}.  This macro informs TPIE
%that it should not rely on the presence of implicit template
%instantiation.  In response to the fact that this macro is set, TPIE
%defines a series of new macros with names of the form {\tt
%  TEMPLATE\_INSTANTIATE\_*}.  
%\index{TEMPLATE_INSTANTIATE_*@{\tt TEMPLATE\_INSTANTIATE\_*}|(}
%Each of these macros can be used to
%actually instantiate some set of functions and/or classes that TPIE
%needs to provide a given operation.  These macros should be used at
%the end of your source file in order to perform the proper
%instantiations.

%The {\tt TEMPLATE\_INSTANTIATE\_*} macros likely to be needed by TPIE
%programmers are as follows:
%\begin{description}
%\item[{\tt TEMPLATE\_INSTANTIATE\_STREAMS(T)}] Instantiate AMI and
%  BTE level streams of objects of type {\tt T}.  If your
%  application uses streams of several types, this macro must be called
%  once for each of them.
%\item[{\tt TEMPLATE\_INSTANTIATE\_ISTREAM(T)}]
%\item[{\tt TEMPLATE\_INSTANTIATE\_OSTREAM(T)}] Instantiate ASCII
%  input and output scan management objects for the type {\tt T}.
%  See Section~\ref{sec:ascii-io} for details on these objects.
%  \index{scanning!ASCII I/O}
%\item[{\tt TEMPLATE\_INSTANTIATE\_AMI\_MERGE}] Instantiate merging entry
%  points for streams of objects of type {\tt T}.  Merging is described
%  in Section~\ref{sec:merging}.
%\item[{\tt TEMPLATE\_INSTANTIATE\_SORT\_OP(T)}]
%\item[{\tt TEMPLATE\_INSTANTIATE\_SORT\_CMP(T)}]
%\item[{\tt TEMPLATE\_INSTANTIATE\_SORT\_OBJ(T)}] Instantiate
%  respectively operator, comparison function, and comparison object
%  based sorting of objects of type {\tt T}.  See
%  Section~\ref{sec:cmp-sorting} for details on these types of sorting.
%\item[{\tt TEMPLATE\_INSTANTIATE\_KB\_SORT(T)}] 
%\item[{\tt TEMPLATE\_INSTANTIATE\_KB\_SORT\_KEY(T,K)}] Instantiate key
%  bucket distribution sorting of objects of type {\tt T}.  The latter
%  form uses key {\tt K} for sorting.  Section~\ref{sec:kb-sorting}
%  describes key bucket sorting.
%\item[{\tt TEMPLATE\_INSTANTIATE\_STREAM\_ADD(T)}]
%\item[{\tt TEMPLATE\_INSTANTIATE\_STREAM\_SUB(T)}]
%\item[{\tt TEMPLATE\_INSTANTIATE\_STREAM\_MULT(T)}]
%\item[{\tt TEMPLATE\_INSTANTIATE\_STREAM\_DIV(T)}]
%  Instantiate elementwise arithmetic operations on streams of objects
%  of type {\tt T} as described in Section~\ref{sec:elementwise}.
%\item[{\tt TEMPLATE\_INSTANTIATE\_AMI\_MATRIX}]
%  Instantiate dense matrices of objects of type {\tt T} and the
%  standard operations on them.  Dense
%  matrices are described in
%  Section~\ref{sec:dense-mat}.\index{matrices!dense}
%\item[{\tt TEMPLATE\_INSTANTIATE\_AMI\_SPARSE\_MATRIX}]
%  Instantiate sparse matrices of objects of type {\tt T} and the
%  standard operations on them.  Sparse
%  matrices are described in
%  Section~\ref{sec:dense-mat}.\index{matrices!sparse}
%\end{description}
%\index{TEMPLATE_INSTANTIATE_*@{\tt TEMPLATE\_INSTANTIATE\_*}|)}

%In addition to instantiating functions and classes using the macros
%described above, it is often necessary to explicitly instantiate
%particular instances of AMI entry points for user supplied operation
%management objects.  For example, suppose we declare a scan management
%object class such as
%\begin{verbatim}
%class my_scan_class : AMI_scan_object {
%public:
%    AMI_err initialize(void);
%    AMI_err operate(const int &in1, const int &in2, AMI_SCAN_FLAG *sfin,
%                    float *out, AMI_SCAN_FLAG *sfout); 
%}
%\end{verbatim}
%Then, in order to explicitly instantiate \myverb{AMI_scan()} to use
%objects of this type, we would use the following code:
%\begin{verbatim}
%template AMI_err AMI_scan(AMI_STREAM<int> *, AMI_STREAM<int> *, 
%                          my_scan_class *, AMI_STREAM<float> *); 
%\end{verbatim}
%This instantiates an instance of \myverb{AMI\_scan()} that takes two input
%streams of \myverb{int}s, operates on them with an object of type
%\myverb{my\_scan\_class}, and produces an output stream of \myverb{float}s.  
%Note the correspondence between the types of input and output streams
%and the types of the operands to the \myverb{operate()} member function
%of the class \myverb{my\_scan\_class}.
%\index{templates!instantiation|)}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
